---
title: "Deforestation trend analysis"
author: "Srijan Nayak"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    toc: true
    toc_float: true
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Objective

The objective of this analysis is to try to see the future state of forest
covers across the world. Is it certain that the forest covers in all countries
will disappear very soon? Or, is there hope that efforts in the recent years
for increasing area under forest cover are not in vain, and thus can we see an
increase in forest covers across countries in the coming years?

```{r include=FALSE}
library(broom)
library(dplyr)
library(ggplot2)
library(purrr)
library(forcats)
library(readr)
library(shiny)
library(tidyr)
```

# Reading data

Data used for this analysis is taken from week 15 of the
[TidyTuesday project][tidytuesday-link] for the year 2021,
[Deforestaion][deforesttion-datasets-link].

[tidytuesday-link]: https://github.com/rfordatascience/tidytuesday
[deforesttion-datasets-link]: https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-04-06/readme.md

```{r}
forest <- read_csv("forest.csv", col_types = "ffdd") %>%
  rename(country = entity) %>%
  filter(country != "World") %>%
  select(-code)
forest_area <- read_csv("forest_area.csv", col_types = "ffdd") %>%
  rename(country = entity) %>%
  filter(country != "World") %>%
  select(-code)

attr(forest, "spec") <- NULL
attr(forest_area, "spec") <- NULL
```

# Exploring and modelling with `forest` data

```{r}
head(forest)
```

## Exploring net forest conversion vs. forest area

The `net_forest_conversion` variable is a bit ambiguous in the sense that it's
not clear if it mean forest cover that has been converted for other purposes, or 
is it land that has been converted to forest cover.

The following interactive application shows the change in net forest conversion
and forest area over the years for a selected country, on the left and right
respectively.

```{r, echo=FALSE}
selectInput("country", "Select country", choices = forest$country)

plot_data <- reactive({
  req(input$country)
  
  forest %>%
    filter(country == input$country) %>%
    left_join(forest_area, by = c("country", "year")) %>%
    pivot_longer(net_forest_conversion:forest_area, names_to = "value_type") %>%
    mutate(
      value_type = fct_relevel(value_type, "net_forest_conversion"),
      value_type = fct_recode(
        value_type,
        "Net forest conversion (hectares)" = "net_forest_conversion",
        "Forest area (% of global forest area)" = "forest_area"
      )
    )
})

renderPlot({
  ggplot(plot_data(), aes(year, value)) +
    geom_line() +
    geom_point() +
    facet_wrap(vars(value_type), scales = "free_y", ncol = 2) +
    theme_minimal() +
    labs(x = "Year", y = "Value")
})
```

Looking at the plots for a few countries it seems that net forest conversion is 
inversely related to forest area, meaning that it represents the deforestation 
amount.

## Splitting data

The data set has only a few years of data for each country, therefore randomly
splitting the data might not leave sufficient training data for some countries.
Thus only the most recent observation is taken in the test data set for each
country.

```{r}
forest_train <- forest %>% 
  group_by(country) %>% 
  filter(year != max(year)) %>% 
  ungroup()
forest_test <- forest %>% 
  group_by(country) %>% 
  filter(year == max(year)) %>% 
  ungroup()
```

### Proportions of train and test data

```{r}
round(nrow(forest_train) / nrow(forest), 2)
```

```{r}
round(nrow(forest_test) / nrow(forest), 2)
```

## Modelling

Instead of keeping `country` as another independent variable in the model,
a separate simple linear regression model is created for each country, with
`net_forest_conversion` as dependent variable and `year` as the independent 
variable. The `fit` variable contains the SLR models for each country.

```{r}
forest_lm_fits <- forest_train %>%
  nest(data = -country) %>%
  mutate(
    fit = map(data, ~ lm(net_forest_conversion ~ year, data = .x)),
    glanced = map(fit, glance),
  )
head(forest_lm_fits)
```

Some countries don't have enough data for training which give poor fitted
models. Thus only models with an $\text{Adjusted } R^2 > 0.6$ are kept.

```{r}
forest_lm_fits_good <- forest_lm_fits %>% 
  unnest(cols = glanced) %>% 
  filter(adj.r.squared > 0.6) %>% 
  select(country, fit)
head(forest_lm_fits_good)
```

### Cheching model accuracy

```{r}
forest_lm_fits_good %>%
  left_join(forest_test, by = "country") %>%
  mutate(
    pred_conversion = round(map2_dbl(fit, year, ~ predict(.x, list(year = .y))))
  ) %>% 
  ggplot(aes(net_forest_conversion, pred_conversion)) +
  geom_point() +
  geom_abline(color = "grey", linetype = "dashed") +
  coord_fixed() +
  theme_minimal() +
  labs(
    title = "Model accuracy",
    x = "Actual forest conversion in hectares",
    y = "Predicted forest conversion in hectares"
  )
```

Most models have their predictions close to the $45^\circ$ reference line
passing through the origin when plotted against the actual forest conversion.
Thus most of the filtered models are decent for making future predictions.

### Predicting net forest conversion for particular country

```{r echo=FALSE}
selectInput("nfc_country", "Select country to predict for",
            choices = forest_lm_fits_good$country)
sliderInput("nfc_year", "Choose year to predict for",
            min = 1980, max = 2080, value = 2021, step = 1, sep = "")

nfc_model <- reactive({
  req(input$nfc_country)
  
  lm_list <- forest_lm_fits_good %>%
    filter(country == input$nfc_country) %>% 
    slice_head(n = 1) %>% 
    pull(fit)
  lm_list[[1]]
})

nfc_prediction <- reactive({
  req(input$nfc_year)

  prediction <- predict(nfc_model(), list(year = input$nfc_year))
  paste("Net forest conversion:", round(prediction, 2), "hectares")
})

renderText(nfc_prediction())
```

# Exploring and modelling with `forest_area` data

```{r}
head(forest_area)
```

```{r}
forest_area %>% 
  count(country) %>% 
  arrange(n) %>% 
  head()
```

## Splitting data

Similar to the previous case, due to the small amount of data only the 4 most 
recent observations have been taken in the test data for each country.

```{r}
forest_area_train <- forest_area %>% 
  group_by(country) %>% 
  filter(!(year %in% sort(year, decreasing = TRUE)[1:4])) %>% 
  ungroup()
forest_area_test <- forest_area %>% 
  group_by(country) %>% 
  filter(year %in% sort(year, decreasing = TRUE)[1:4]) %>% 
  ungroup()
```

### Proportions of train and test data

```{r}
round(nrow(forest_area_train) / nrow(forest_area), 2)
```

```{r}
round(nrow(forest_area_test) / nrow(forest_area), 2)
```

## Modelling

Just like before, a separate SLR model is created for each country...

```{r}
forest_area_lm_fits <- forest_area_train %>% 
  nest(data = -country) %>% 
  mutate(
    fit = map(data, ~ lm(forest_area ~ year, data = .x)),
    glanced = map(fit, glance)
  )
head(forest_area_lm_fits)
```

...and only models with an $\text{Adjusted } R^2 > 0.6$ are kept.

```{r}
forest_area_lm_fits_good <- forest_area_lm_fits %>% 
  unnest(cols = glanced) %>% 
  filter(adj.r.squared > 0.6) %>% 
  select(country, fit)
head(forest_area_lm_fits_good)
```

### Cheching model accuracy

```{r}
forest_area_lm_fits_good %>%
  left_join(forest_area_test, by = "country") %>%
  mutate(
    pred_area = round(map2_dbl(fit, year, ~ predict(.x, list(year = .y))))
  ) %>%
  ggplot(aes(forest_area, pred_area)) +
  geom_point() +
  geom_abline(color = "grey", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Model accuracy",
     x = "Actual forest area (% of global forest area)",
     y = "Predicted forest area (% of global forest area)"
  ) +
  coord_fixed()
```

Judging from the plot, the model seems to perform really well when compared with
the actual forest areas from the test data set.

### Predicting proportion of global forest area for particular country

```{r echo=FALSE}
selectInput("fa_country", "Select country to predict for",
            choices = forest_area_lm_fits_good$country)
sliderInput("fa_year", "Choose year to predict for",
            min = 1980, max = 2080, value = 2021, step = 1, sep = "")

fa_model <- reactive({
  req(input$fa_country)
  
  lm_list <- forest_area_lm_fits_good %>%
    filter(country == input$fa_country) %>% 
    slice_head(n = 1) %>% 
    pull(fit)
  lm_list[[1]]
})

fa_prediction <- reactive({
  req(input$fa_year)

  prediction <- predict(fa_model(), list(year = input$fa_year))
  paste("Forest area:", round(prediction, 5), "%")
})

renderText(fa_prediction())
```

# Conclusion

Just looking at the predictions, it seems that situation may not be as bad as we 
already believe. A lot of countries seem to be following a trend of lower net 
forest conversions throughout the years and almost all countries seem to be 
following an uptrend for larger forest areas.

It should be noted that this analysis was done with very limited data and a lot
of factors have not been accounted for. Thus even though the predictions seem
to be showing a positive future, we should not let our guard down and actively
try to battle deforestation in our own individual ways.
