---
title: "Deforestation trend analysis"
author: "Srijan Nayak"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    toc: true
    toc_float: true
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r include=FALSE}
library(broom)
library(dplyr)
library(ggplot2)
library(purrr)
library(forcats)
library(readr)
library(shiny)
library(tidyr)
```

# Reading data

Data used for this analysis is taken from week 15 of the
[TidyTuesday project][tidytuesday-link] for the year 2021,
[Deforestaion][deforesttion-datasets-link].

[tidytuesday-link]: https://github.com/rfordatascience/tidytuesday
[deforesttion-datasets-link]: https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-04-06/readme.md

```{r}
forest <- read_csv("forest.csv", col_types = "ffdd") %>%
  rename(country = entity) %>%
  filter(country != "World") %>%
  select(-code)
forest_area <- read_csv("forest_area.csv", col_types = "ffdd") %>%
  rename(country = entity) %>%
  filter(country != "World") %>%
  select(-code)

attr(forest, "spec") <- NULL
attr(forest_area, "spec") <- NULL
```

# Exploring and modelling with `forest` data

```{r}
head(forest)
```

## Exploring net forest conversion vs. forest area

```{r, echo=FALSE}
selectInput("country", "Select country", choices = forest$country)

plot_data <- reactive({
  req(input$country)
  
  forest %>%
    filter(country == input$country) %>%
    left_join(forest_area, by = c("country", "year")) %>%
    pivot_longer(net_forest_conversion:forest_area, names_to = "value_type") %>%
    mutate(
      value_type = fct_relevel(value_type, "net_forest_conversion"),
      value_type = fct_recode(
        value_type,
        "Net forest conversion (hectares)" = "net_forest_conversion",
        "Forest area (% of global forest area)" = "forest_area"
      )
    )
})

renderPlot({
  ggplot(plot_data(), aes(year, value)) +
    geom_line() +
    geom_point() +
    facet_wrap(vars(value_type), scales = "free_y", ncol = 2) +
    theme_minimal() +
    labs(x = "Year", y = "Value")
})
```

## Splitting data

```{r}
forest_train <- forest %>% 
  group_by(country) %>% 
  filter(year != max(year)) %>% 
  ungroup()
forest_test <- forest %>% 
  group_by(country) %>% 
  filter(year == max(year)) %>% 
  ungroup()
```

### Proportions of train and test data

```{r}
round(nrow(forest_train) / nrow(forest), 2)
```

```{r}
round(nrow(forest_test) / nrow(forest), 2)
```

## Modelling

```{r}
forest_lm_fits <- forest_train %>%
  nest(data = -country) %>%
  mutate(
    fit = map(data, ~ lm(net_forest_conversion ~ year, data = .x)),
    glanced = map(fit, glance),
  )
head(forest_lm_fits)
```

```{r}
forest_lm_fits_good <- forest_lm_fits %>% 
  unnest(cols = glanced) %>% 
  filter(adj.r.squared > 0.6) %>% 
  select(country, fit)
head(forest_lm_fits_good)
```

### Cheching model accuracy

```{r}
forest_lm_fits_good %>%
  filter(country != "World") %>% 
  left_join(forest_test, by = "country") %>%
  mutate(
    pred_conversion = round(map2_dbl(fit, year, ~ predict(.x, list(year = .y))))
  ) %>% 
  ggplot(aes(net_forest_conversion, pred_conversion)) +
  geom_point() +
  geom_abline(color = "grey", linetype = "dashed") +
  coord_fixed() +
  theme_minimal() +
  labs(
    title = "Model accuracy",
    subtitle = "Predicting for year 2015",
    x = "Actual forest conversion in hectares",
    y = "Predicted forest conversion in hectares"
  )
```

### Predicting net forest conversion for particular country

```{r echo=FALSE}
selectInput("nfc_country", "Select country to predict for",
            choices = forest_lm_fits_good$country)
sliderInput("nfc_year", "Choose year to predict for",
            min = 1980, max = 2080, value = 2021, step = 1, sep = "")

nfc_model <- reactive({
  req(input$nfc_country)
  
  lm_list <- forest_lm_fits_good %>%
    filter(country == input$nfc_country) %>% 
    slice_head(n = 1) %>% 
    pull(fit)
  lm_list[[1]]
})

nfc_prediction <- reactive({
  req(input$nfc_year)

  prediction <- predict(nfc_model(), list(year = input$nfc_year))
  paste("Net forest conversion:", round(prediction, 2), "hectares")
})

renderText(nfc_prediction())
```

# Exploring and modelling with `forest_area` data

```{r}
head(forest_area)
```

```{r}
forest_area %>% 
  count(country) %>% 
  arrange(n) %>% 
  head()
```

## Splitting data

```{r}
forest_area_train <- forest_area %>% 
  group_by(country) %>% 
  filter(!(year %in% sort(year, decreasing = TRUE)[1:4])) %>% 
  ungroup()
forest_area_test <- forest_area %>% 
  group_by(country) %>% 
  filter(year %in% sort(year, decreasing = TRUE)[1:4]) %>% 
  ungroup()
```

### Proportions of train and test data

```{r}
round(nrow(forest_area_train) / nrow(forest_area), 2)
```

```{r}
round(nrow(forest_area_test) / nrow(forest_area), 2)
```

## Modelling

```{r}
forest_area_lm_fits <- forest_area_train %>% 
  nest(data = -country) %>% 
  mutate(
    fit = map(data, ~ lm(forest_area ~ year, data = .x)),
    glanced = map(fit, glance)
  )
head(forest_area_lm_fits)
```

```{r}
forest_area_lm_fits_good <- forest_area_lm_fits %>% 
  unnest(cols = glanced) %>% 
  filter(adj.r.squared > 0.6) %>% 
  select(country, fit)
head(forest_area_lm_fits_good)
```

### Cheching model accuracy

```{r}
forest_area_lm_fits_good %>%
  left_join(forest_area_test, by = "country") %>%
  mutate(
    pred_area = round(map2_dbl(fit, year, ~ predict(.x, list(year = .y))))
  ) %>%
  ggplot(aes(forest_area, pred_area)) +
  geom_point() +
  geom_abline(color = "grey", linetype = "dashed") +
  facet_wrap(vars(year), ncol = 2) +
  theme_minimal() +
  labs(
    title = "Model accuracy",
     x = "Actual forest area (% of global forest area)",
     y = "Predicted forest area (% of global forest area)"
  ) +
  coord_fixed()
```

### Predicting proportion of global forest area for particular country

```{r echo=FALSE}
selectInput("fa_country", "Select country to predict for",
            choices = forest_area_lm_fits_good$country)
sliderInput("fa_year", "Choose year to predict for",
            min = 1980, max = 2080, value = 2021, step = 1, sep = "")

fa_model <- reactive({
  req(input$fa_country)
  
  lm_list <- forest_area_lm_fits_good %>%
    filter(country == input$fa_country) %>% 
    slice_head(n = 1) %>% 
    pull(fit)
  lm_list[[1]]
})

fa_prediction <- reactive({
  req(input$fa_year)

  prediction <- predict(fa_model(), list(year = input$fa_year))
  paste("Forest area:", round(prediction, 2), "%")
})

renderText(fa_prediction())
```
